---
title: "Proportions Notebook"
output: 
  md_document:
  html_document:
    highlight: tango
    theme: cerulean
    toc: yes
    toc_depth: 4
    toc_float: yes
---

This lecture was taken from Statistical Analysis and Reporting in R, Jacob O. Wobbrock, Ph.D.
available here: http://depts.washington.edu/acelab/proj/Rstats/index.html

 Proportions.R
(Tests of Proportion & Association)

One Sample


#### Binomial test

When to use this test? 

*Notes taken from here:https://www.youtube.com/watch?v=WOoS7nVkfDk by Matthew E.Clapham

*Type of Data*: Categorial, only have two categories in a single sample

*Purpose*: To test if Category has an expected count(Goodness of fit).

The *goodness of fit* of a statistical model describes how well it fits a set of observations. Measures of goodness of fit typically summarize the discrepancy between observed values and the values expected under the model in question.


```{r setup}

knitr::opts_chunk$set(echo = TRUE)
# pull in the csv
df <- read.csv("data/Proportions/0F0LBs_binomial.csv")
df
```
##### Create factors
Subject id is nominal and Y is categorical

```{r factors}

df$S = factor(df$S) # Subject id is nominal (unused)
df$Y = factor(df$Y) # Y is an outcome of exactly two categories

```
Column Y is made up of two values: `r unique(df$Y)`
```{r xtabs}
xt = xtabs( ~ Y, data=df) # make counts

```
`r xt`

binom.test 

Exact Binomial Test
Performs an exact test of a simple null hypothesis about the probability of success in a Bernoulli experiment.
https://www.rdocumentation.org/packages/stats/versions/3.1.1/topics/binom.test

```{r binomial_test}
binom.test(xt, p=0.5, alternative="two.sided")
```


#### Multinomial test
*Type of Data*: Categorial, 3+ categories in a single sample

*Purpose*: To test if Categories match with expected count(Goodness of fit).

We will use the XNomial library
https://cran.r-project.org/web/packages/XNomial/index.html

```{r Multinomial_library}
# df is a long-format data table w/columns for subject (S) and N-category outcome (Y)
if (!require("XNomial")) install.packages("XNomial");  
library(XNomial) # for xmulti
```

```{r Multinomial_dataset}
df_multi <- read.csv("data/Proportions/0F0LBs_multinomial.csv")
```
```{r Multinomial_factor}
df_multi$S = factor(df$S) # Subject id is nominal (unused)
df$Y = factor(df$Y) # Y is an outcome of ≥2 categories
```
```{r Multinomial_tabs}
xt = xtabs( ~ Y, data=df) # make counts
```
```{r Multinomial_test}
xmulti(xt, rep(1/length(xt), length(xt)), statName="Prob")
```