---
title: "Proportions Notebook"
output: 
  html_document:
    highlight: tango
    theme: cerulean
    toc: yes
    toc_depth: 4
    toc_float: yes
  md_document:
    toc: yes
---

This code was taken from Statistical Analysis and Reporting in R, Jacob O. Wobbrock, Ph.D.
available here: http://depts.washington.edu/acelab/proj/Rstats/index.html

 Proportions.R
(Tests of Proportion & Association)

One Sample


### Binomial test

When to use this test? 

*Notes taken from here:https://www.youtube.com/watch?v=WOoS7nVkfDk by Matthew E.Clapham

**Type of Data**: Categorial, only have two categories in a single sample

**Purpose**: To test if Category has an expected count(Goodness of fit).

The **goodness of fit** of a statistical model describes how well it fits a set of observations. Measures of goodness of fit typically summarize the discrepancy between observed values and the values expected under the model in question.


```{r setup}
# pull in the csv
df <- read.csv("data/Proportions/0F0LBs_binomial.csv")
head(df)
```
##### Create factors
Subject id is nominal and Y is categorical

```{r factors}

df$S = factor(df$S) # Subject id is nominal (unused)
df$Y = factor(df$Y) # Y is an outcome of exactly two categories

```
Column Y is made up of two values: `r unique(df$Y)`
```{r xtabs}
xt = xtabs( ~ Y, data=df) # make counts

```
`r xt`

binom.test 

**Exact Binomial Test**
Performs an exact test of a simple null hypothesis about the probability of success in a Bernoulli experiment.
https://www.rdocumentation.org/packages/stats/versions/3.1.1/topics/binom.test

```{r binomial_test}
binom.test(xt, p=0.5, alternative="two.sided")
```


### Multinomial test
**Type of Data**: Categorial, 3+ categories in a single sample

**Purpose**: To test if Categories match with expected count(Goodness of fit).

We will use the XNomial library
https://cran.r-project.org/web/packages/XNomial/index.html


```{r Multinomial_library}
# df is a long-format data table w/columns for subject (S) and N-category outcome (Y)
if (!require("XNomial")) install.packages("XNomial");  
library(XNomial) # for xmulti

df_multi <- read.csv("data/Proportions/0F0LBs_multinomial.csv")
head(df_multi) 
```
```{r Multinomial_factor}
df_multi$S = factor(df_multi$S) # Subject id is nominal (unused)
df_multi$Y = factor(df_multi$Y) # Y is an outcome of ≥2 categories
xt = xtabs( ~ Y, data=df_multi) # make counts
xt # 3 categorical variables
```
```{r Multinomial_test}
xmulti(xt, rep(1/length(xt), length(xt)), statName="Prob")
```
```{r Multinomial_test2}
# the following gives the equivalent result
if (!require("RVAideMemoire")) install.packages("RVAideMemoire");  
library(RVAideMemoire) # for multinomial.test

multinomial.test(df_multi$Y)
```
Multinomial post hoc test
```{r Multinomial_post_hoc}
# xt is a table of counts for each category of Y
library(RVAideMemoire) # for multinomial.multcomp
multinomial.multcomp(xt, p.method="holm") # xt shows levels
```
### Chi-squared test 
*notes taken from http://www.sthda.com/english/wiki/chi-square-goodness-of-fit-test-in-r

The **chi-square goodness of fit test** is used to compare the observed distribution to an expected distribution, in a situation where we have two or more categories in a discrete data. In other words, it compares multiple observed proportions to expected probabilities.

```{r Multinomial_chi_data}
## One-Sample Chi-Squared test
# df is a long-format data table w/columns for subject (S) and N-category outcome (Y)
df <- read.csv("data/Proportions/0F0LBs_multinomial.csv")
head(df)
```

```{r Multinomial_tab}
df$S = factor(df$S) # Subject id is nominal (unused)
df$Y = factor(df$Y) # Y is an outcome from two or more categories
xt = xtabs( ~ Y, data=df) # make counts
xt
```

```{r Multinomial_chi}
result <-chisq.test(xt)
result 
```
The p-value of the test is 7.869e-05, which is less than the significiance level alpha =0.05. We can conclude that the outcomes are not evenly distributed p-value of 7.869e-05.
```{r chi_expected}
# access of the expected value
result$expected
```
```{r chi_p_value}
# access of the expected value
result$p.value
```

```{r chi_mult}
## Chi-Squared post hoc test
# xt is a table of counts for each category of Y
library(RVAideMemoire) # for chisq.multcomp
chisq.multcomp(xt, p.method="holm") # xt shows levels
# for the Chi-Squared values, use qchisq(1-p, df=1), where p is the pairwise p-value.
```

**One-sample post hoc tests**

A different kind of post hoc test for one sample. For Y's response categories (x,y,z), test each proportion against chance.


Two Samples

### Fisher's exact test
*notes from https://www.youtube.com/watch?v=WTDWk4eJIw0 by 
Matthew E. Clapham

Method for checking for independance, generally in small samples and small contingency table.

**Types of data**: Categorical, 2x2 contingency table. Technically both row and column marginals should be fixed prior to data collection.

**Purpose**: To test for assocation between the two counts. 

**null hypothesis**: Category frequencies are independent of one another between the samples (i.e, there is no association amoung categories)


```{r fishers_df}
# df is a long-format data table w/subject (S), categorical factor (X) and outcome (Y)
df <- read.csv("data/Proportions/1F2LBs_multinomial.csv")
head(df)
```

```{r fishers_df_factor}
df$S = factor(df$S) # Subject id is nominal (unused)
df$X = factor(df$X) # X is a factor of m ≥ 2 levels
df$Y = factor(df$Y) # Y is an outcome of n ≥ 2 categories
```

```{r fishers_levels}
unique(df$X)
unique(df$Y)

```

```{r fishers_tabs}
xt = xtabs( ~ X + Y, data=df) # make m × n crosstabs
xt
```


```{r fishers_test}
res <- fisher.test(xt)
res
```

```{r fishers_post}
## Fisher's post hoc test
# xt is an m × n crosstabs with categories X and Y
library(RVAideMemoire) # for fisher.multcomp
fisher.multcomp(xt, p.method="holm")
```


### G-test
```{r g_test_data}
# df is a long-format data table w/subject (S), categorical factor (X) and outcome (Y)
library(RVAideMemoire) # for G.test
df <- read.csv("data/Proportions/1F2LBs_multinomial.csv")
head(df)
```
```{r g_test_factor}
df$S = factor(df$S) # Subject id is nominal (unused)
df$X = factor(df$X) # X is a factor of m ≥ 2 levels
df$Y = factor(df$Y) # Y is an outcome of n ≥ 2 categories
```
```{r g_test_tabs}
xt = xtabs( ~ X + Y, data=df) # make m × n crosstabs
```
```{r g_test}
G.test(xt)
```
References: 

http://www.sthda.com/english/wiki/comparing-proportions-in-r
http://www.sthda.com/english/wiki/chi-square-goodness-of-fit-test-in-r#what-is-chi-square-goodness-of-fit-test
